---
title: "Homework: HbA1c Control Trends — Clinics B & D, and a Simple ITS v2"
format:
  html:
    toc: true
    code-tools: true
    embed-resources: true
execute:
  echo: true
  warning: true
---

## Setup

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(lubridate)
library(qicharts2)
```

## Load the data

```{r}
#| label: load-data
#| message: false

encounters <- read_csv("data/encounters.csv")
labs <- read_csv("data/labs.csv")
lookup <- read_csv("data/provider_lookup.csv")
```

Same three files as class. This time you're finishing Clinics B and D,
which have two additional problems A and C didn't.

## Part 1: Build the measure

### Parse dates (same as class)

```{r}
#| label: parse-dates

# TODO: reuse your mixed-date parser from class on encounter_date and
# result_date.

parse_mixed_date <- function(x) {
  d_iso <- suppressWarnings(as.Date(x, format = "%Y-%m-%d"))
  d_us  <- suppressWarnings(as.Date(x, format = "%m/%d/%Y"))
  coalesce(d_iso, d_us)
}

encounters <- encounters %>% mutate(encounter_date = parse_mixed_date(encounter_date))
labs <- labs %>% mutate(result_date = parse_mixed_date(result_date))

sum(is.na(encounters$encounter_date))
sum(is.na(labs$result_date))
```

### Task 1: standardize the join key, then join

```{r}
#| label: standardize-and-join

# TODO: first, try the plain join from class:
#   encounters %>% left_join(lookup, by = "provider_id")
# and check how many rows come back with a missing clinic_id. Then look
# at a sample of the unmatched provider_id values -- what's inconsistent
# about them relative to provider_lookup.csv?
#
# TODO: create a standardized version of provider_id on BOTH tables
# (case + whitespace), then join on the standardized version. Confirm
# zero unmatched rows afterward.

encounters <- read_csv("data/encounters.csv")

encounters <- encounters %>%
  left_join(lookup, by = "provider_id")

encounters %>%
  filter(is.na(clinic_id)) %>%
  select(provider_id) %>%
  head(10)

encounters <- encounters %>%
  mutate(provider_id_clean = toupper(trimws(provider_id)))

lookup <- lookup %>%
  mutate(provider_id_clean = toupper(trimws(provider_id)))

encounters <- encounters %>%
  select(-clinic_id) %>%
  left_join(
    lookup %>% select(provider_id_clean, clinic_id),
    by = "provider_id_clean"
  )

sum(is.na(encounters$clinic_id))
```

### Build the denominator

```{r}
#| label: denominator

# TODO: same logic as class,
# filter to Clinic B and Clinic D.
# set dates to month level
# reduce to unique patients per clinic per month rows
# count the number for rows

encounters <- encounters %>% 
  mutate(encounter_date = parse_mixed_date(encounter_date))

labs <- labs %>% 
  mutate(result_date = parse_mixed_date(result_date))

denom <- encounters %>%
  filter(clinic_id %in% c("Clinic B", "Clinic D")) %>%
  mutate(month = floor_date(encounter_date, "month")) %>%
  distinct(clinic_id, month, patient_id) %>%
  count(clinic_id, month, name = "denominator")

head(denom)
```

### Clean the labs and handle Task 2: implausible values

```{r}
#| label: clean-labs-filter

# TODO: filter to test_name == "HbA1c", convert result_value to numeric,
# drop non-numeric/missing entries, remove exact duplicates -- same logic as
# class.

hba1c_clean <- labs %>%
  filter(test_name == "HbA1c") %>%
  distinct(patient_id, result_date, test_name, result_value) %>%
  mutate(result_value = suppressWarnings(as.numeric(result_value))) %>%
  filter(!is.na(result_value)) %>%
  mutate(month = floor_date(result_date, "month"))
```

```{r}
#| label: plausible-range


# TODO (Task 2): before computing the numerator, add a plausible-range
# filter on the numeric value (a reasonable range for this exercise is
# roughly 4-15). Report how many rows this excludes. Also convert dates to month. 
sum(hba1c_clean$result_value < 4 | hba1c_clean$result_value > 15)

# This excludes 18 rows.

hba1c_clean <- hba1c_clean %>%
  filter(result_value >= 4, result_value <= 15)

sum(hba1c_clean$result_value < 4 | hba1c_clean$result_value > 15)
```

### Build the numerator and assemble the measure table

```{r}
#| label: numerator-patients

# TODO: same join-back-to-clinic-month logic as class
# First, identify unique patient visits in a month.

patient_clinic_month <- encounters %>%
  filter(clinic_id %in% c("Clinic B", "Clinic D")) %>%
  mutate(month = floor_date(encounter_date, "month")) %>%
  distinct(clinic_id, month, patient_id)

head(patient_clinic_month)
```

```{r}
#| label: numerator-labs
# Finish generating a numerator - # of patients with controlled values (under 8)
# Filter to controlled lab values, join to the patient-by month file above.
# Identify unique patients per month per clinic
# Count how many patients per month had controlled lab values. This is the numerator.
# 
numer <- hba1c_clean %>%
  filter(result_value < 8) %>%
  inner_join(patient_clinic_month, by = c("patient_id", "month")) %>%
  distinct(clinic_id, month, patient_id) %>%
  count(clinic_id, month, name = "numerator")

head(numer)
```

```{r}
#| label: numerator-assembly
# Join numerator and denominator together
measure <- denom %>%
  left_join(numer, by = c("clinic_id", "month")) %>%
  mutate(numerator = coalesce(numerator, 0L))

head(measure)

```


## Part 2: Control charts

```{r}
#| label: chart-b

# TODO: qic() with chart = "p" for Clinic B. Store the result.

q_b <- qic(
  x = month, y = numerator, n = denominator,
  data = measure %>% filter(clinic_id == "Clinic B") %>% arrange(month),
  chart = "p", title = "Clinic B"
)
q_b

any(q_b$data$runs.signal, na.rm = TRUE)
```

*Clinic B interpretation:*

> Runs signal present. Most points are below center line from the start in 2023 to a early 2025. Then it is above the center line from early 2025 to 2026 but its still fluctuating all throughout the months. Plausible cause for the fluctuations could be a change in patient management or a significant workflow change. 
[3-5 sentences: signal present? sigma or runs? plausible cause? next
> step?]
>
> **One-line status:** Clinic B - runs signal present, look into changing overall workflow process

```{r}
#| label: chart-d

# TODO: qic() with chart = "p" for Clinic D.
q_d <- qic(
  x = month, y = numerator, n = denominator,
  data = measure %>% filter(clinic_id == "Clinic D") %>% arrange(month),
  chart = "p", title = "Clinic D"
)
q_d

any(q_d$data$runs.signal, na.rm = TRUE)
```

*Clinic D interpretation:*

> Runs signal present. Most points are above center line from the start in 2023 to a little after 2023. Then it is below the center line until a little after 2024. From then to 2026, most points are above the center line, but it is still fluctuating throughout the months. Plausible cause for the fluctuations could be a change in patient management or a significant workflow change.

>
> **One-line status:** Clinic D - runs signal present, look into changing overall workflow process

## Part 3: A simple interrupted time series (Clinic B)

```{r}
#| label: its-setup

# TODO: build a data frame for Clinic B with:
#   - rate = numerator / denominator
#   - t = a simple 1, 2, 3, ... month counter
#   - post = 0/1 indicator for month >= 2025-03-01
#   - t_post = post * (a counter that starts at 1 on the intervention
#     month and counts up from there, 0 before it)
sits_b <- measure %>%
  filter(clinic_id == "Clinic B") %>%
  arrange(month) %>%
  mutate(
    rate = numerator / denominator,
    t = 1:n(),
    post = ifelse(month >= as.Date("2025-03-01"), 1, 0),
    t_post = ifelse(post == 1, t - which(month == as.Date("2025-03-01")) + 1, 0)
  )

head(sits_b)
```

```{r}
#| label: its-model

# TODO: fit lm(rate ~ t + post, data = ...) and look at
# summary(). Extract/interpret the post and t_post coefficients.
sits_model <- lm(rate ~ t + post + t_post, data = sits_b)

summary(sits_model)
```

*Interpretation:*

> [What does the `post` coefficient mean, and is it significant? What
> does `t_post` mean, and is it significant? Does this agree with what
> the control chart's signal already told you? Explain, in your own
> words, what each method is actually testing and why they might (or
> might not) agree.]
The post coefficient is 0.163 and is significant because p = 0.004. This means that the HbA1c control rate increased after March 2025. The t_post coefficient is 0.001 and is not significant because p = 0.922. This means that there was no significant change in the slope. This matches the control chart because the p chart showed a shift from mostly below the center line to mostly above it after 2025.
